Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-30 docs/manuscript/paper.tex
plot updates for smaller strains

Since we are not setting priors on $\gamma^{(n)}$ or $d^{(n)}$ this reduces the dimensionality of the parameter space to $7 + 3N$. We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f. Equation \ref{eq:psrparams}. Explicitly, 
\begin{equation}
	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ .
\end{equation}
All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
\begin{table*}
	\centering
%	\resizebox{\textwidth}{!}{%
%	\renewcommand{\arraystretch}{1.0} % Default value: 1
	\begin{tabular}{lccll}
		\toprule
		&Parameter & Injected Values & Units & Prior  \\
		\hline
		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
	  & $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
	  & $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
	  & $h_0$            & $10^{-12}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
	  & $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
		\hline
		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
		\bottomrule
	\end{tabular}
	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter. The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}. The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
	\label{tab:parameters_and_priors}
\end{table*}

@@

 The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$. The histograms shown in the Figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value. The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours. It is evident that for this representative example we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for the majority of the parameters. The clear exception is $\iota$ which exhibits a notable bias, and lies outside the with the 90\% credible interval, with the posterior median shifted from the injected value by $\sim 0.3$ radians. For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e. a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation. We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}. If we consider the median values of the one-dimensional posteriors we can see that for some parameters the median is almost identically equal to the injected value  (e.g. $\Omega, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g. $\psi, \alpha$), whilst still being contained within the credible interval. This potential bias will also be investigated alongside the bias in $\iota$ in Section \ref{sec:multiple_noise} and Section \ref{sec:bias}. Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work. We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. We are able to accurately estimate each parameter of interest. We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{12}} = h_0 \times 10^{12}$.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f. Figure \ref{fig:likelihood_surface}.} 
	\label{fig:corner_plot_2}
\end{figure*}

@@

To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive of similarity between probability distributions.

A full review of the Wasserstein distance which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein}

In order to quantitatively define `similar'  we use . The WD is a metric which quantifies the distance between two probability distributions 
as the optimal cost, with respect to some cost function $c(x,y)$,  of transporting a probability mass from position $x$ to position $y$. The $p$-Wasserstein distance between two distributions $\mu, \nu$ is
\begin{eqnarray}
	W_p(\mu,\nu)= \left( \inf_{\gamma \in \Gamma(\mu, \nu)}  \int c(x,y)^p d \gamma (x,y)\right)^{1/p}
\end{eqnarray}
where $\gamma(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $c(x,y)$ is the cost function.

In order to compare the control and competitor ensemble probability distributions we adopt as our key metric the Wasserstein distance (WD). The WD intuitively defines a distance between s

We also take $p=1$ and $c(x,y) = |x-y|$ such that the WD values quoted in this work are given by
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\gamma \in \Gamma(\mu, \nu)} \int |x-y| d \gamma (x,y)
\end{eqnarray}
From the perspective of climate modelling, the WD has a nice physical interpretation given by the Monge-Kantorovich duality:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
\end{eqnarray}
i.e. the difference in the expected value of some random variable between two distributions is no bigger than the WD value itself. More explicitly, if the WD for surface temperature between two distributions is 1K, then this tells us that the difference in the expected temperature between the distributions is no greater than 1K. For a full review of the Wasserstein distance, including favourable properties and comparison with other metrics we refer the reader to the Appendix of Pax21.

@@

\textcolor{red}{Wasserstein intro and discussion here}

@@

\subsection{Bias by dropping pulsar terms}\label{sec:bias}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and the Earth terms model for Earth terms only synthetic data (green line), with respect to the inclination $\iota$, over the prior space. All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}. The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model. For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$). For the pulsar-terms model the location of the maxima and injected value do coincide. 10 noise realisations are plotted but the solutions are completely overlaid.}
	\label{fig:likelihood_surface_iota}
\end{figure}
Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g. the median of the one-dimensional marginalised posterior) is biased away from the true injection value. This effect is most severe for $\iota$ and $h_0$ but also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$. In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}. \newline

@@

\appendix

\section{Kalman recursion equations} \label{sec:kalman}
The linear Kalman filter operate on temporally discrete, noisy measurements $\boldsymbol{Y}_k$ which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$ via a linear transformation:
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter takes the underlying states to evolve according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the input or control matrix. and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$. \newline

@@

This paper is useful: \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}
\section{Alternative version of Fig 1}

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{images/Kalman_example_both}
	\caption{\textcolor{red}{TK: For consideration. Do we prefer this figure to Figure \ref{fig:four figures}}?}
	\label{fig:kalmanexampleboth}
\end{figure*}

@@

\section{References}
\label{sec:ref_list}

>>>>

Since we are not setting priors on $\gamma^{(n)}$ or $d^{(n)}$ this reduces the dimensionality of the parameter space to $7 + 3N$. We use the notation $\boldsymbol{\theta}_{\rm psr, reduced}$ to refer to the reduced parameter space, c.f. Equation \ref{eq:psrparams}. Explicitly, 
\begin{equation}
	\boldsymbol{\theta}_{\rm psr, reduced} = \left \{ f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\sigma^{(n)}\right\}_{1\leq n \leq N} \ .
\end{equation}
All the injected static parameters and their corresponding priors for this representative example are summarised in Table \ref{tab:parameters_and_priors}.
\begin{table*}
	\centering
%	\resizebox{\textwidth}{!}{%
%	\renewcommand{\arraystretch}{1.0} % Default value: 1
	\begin{tabular}{lccll}
		\toprule
		&Parameter & Injected Values & Units & Prior  \\
		\hline
		\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
	  & $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
	  & $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
	  & $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
	  & $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
	  & $h_0$            & $5 \times 10^{-15}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
	  & $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
		\hline
		\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
		& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
		&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
		& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-5/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
		& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
		\bottomrule
	\end{tabular}
	\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter. The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}. The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
	\label{tab:parameters_and_priors}
\end{table*}

@@

 The results are shown in Figure \ref{fig:corner_plot_1} for the seven parameters of  $\boldsymbol{\theta}_{\rm gw}$. The histograms shown in the Figure are the one-dimensional posteriors for each parameter, marginalized over the set of all other parameters, where the dashed vertical blue lines show the 0.16 and 0.84 quantiles and the solid orange line the true injection value. The two-dimensional contours denote the (0.5, 1, 1.5, 2)-sigma equivalent contours. It is evident that for this representative example with strain $h_0 = 5 \times 10^{-15}$ that we are able to estimate all the static parameters of the system using our method, with the injected value contained within the 90\% credible interval for all of the parameters.

 If we consider the median values of the one-dimensional posteriors we can see that for some parameters the median is almost identically equal to the injected value (e.g. $\Omega)

 The clear exception is $\iota$ which exhibits a notable bias, and lies outside the with the 90\% credible interval, with the posterior median shifted from the injected value by $\sim 0.3$ radians.

 For a single realisation of the noise it is not clear if this discrepancy is a systematic effect, i.e. a bias in the parameter estimate, or else just a random variation specific to this particular noise realisation. We explore this further in Section \ref{sec:multiple_noise} and Section \ref{sec:bias} and show that this is a bias that results from dropping the pulsar terms, similar to the bias reported in \cite{Zhupulsarterms}.

  $, \Phi_0, \delta$) whereas for others the median has a slight discrepancy from the injected value (e.g. $\psi, \alpha$), whilst still being contained within the credible interval.

  This potential bias will also be investigated alongside the bias in $\iota$ in Section \ref{sec:multiple_noise} and Section \ref{sec:bias}.

 Similar results are derived for the the 3$N$ parameters of $\boldsymbol{\theta}_{\rm psr, reduced}$ and the parameters are generally recovered unambiguously, but we do not display the resulting posterior distributions here since the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this work. We are able to simultaneously estimate both $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr, reduced}$.

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/small_h_posterior_10}
	\caption{Posterior distribution for the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system described in Table \ref{tab:parameters_and_priors}, for a single realisation of the system noise. The vertical orange lines indicate the true injected values. The contours in the 2D histograms denote the (0.5, 1, 1.5, 2)-$\sigma$ levels. We are able to accurately estimate each parameter of interest. We pot the scaled variables $\Omega_{\rm nHz} = \Omega \times 10^{9}$ Hz and $h_{0, \times 10^{15}} = h_0 \times 10^{15}$. Note that the $x$-axis scaling of the one-dimensional posteriors does not span the total prior space.}
	\label{fig:corner_plot_1}
\end{figure*}

@@

\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
%\begin{figure*}
%	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_1000}
%	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f. Figure \ref{fig:likelihood_surface}.} 
%	\label{fig:corner_plot_2}
%\end{figure*}

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/stacked_GW_plot_small_h_definitive2}
	\caption{As Figure \ref{fig:corner_plot_1} but for 9 realisations of the noise processes. The inferred posterior distributions show strong agreement between the different noise realisations for the majority of static parameters. The exceptions are $\iota$ and $h$ which exhibit a larger degree of variance due to the weak identifiability between these two parameters c.f. Figure \ref{fig:likelihood_surface}.} 
	\label{fig:corner_plot_2}
\end{figure*}

@@

To this end we use the the Wasserstein distance \citep[WD,][]{Wasserstein,Villani2009} from optimal transport theory which defines an intuitive notion of similarity between probability distributions. The WD is the cost of an optimal strategy for moving probability mass between two distributions from position $x$ to position $y$, with respect to some cost function $c(x,y)$. Throughout this work we use the first moment of the Wasserstein distance which for one-dimensional distributions $\mu$ and  $\nu$ is defined
\begin{eqnarray}
	W_1(\mu,\nu)=  \inf_{\lambda \in \Gamma(\mu, \nu)} \int |x-y| d \lambda (x,y)
\end{eqnarray}
where $\lambda(x,y)$ is the transport plan, $\Gamma(\mu, \nu)$ is the set of all couplings of $\mu$ and $\nu$ and $ |x-y|$ defines the cost function. The WD has an additional useful physical interpretation given by the Monge-Kantorovich duality:
\begin{eqnarray}
	| \boldsymbol{E}(X_{\mu} )-\boldsymbol{E}(Y_{\nu} ) | \leq W_1(\mu, \nu) \label{eq:WDdefn}
\end{eqnarray}
where $X_{\mu}$ and $Y_{\nu}$ are random variables of the distributions $\mu$ and $\nu$ respectively, and $\boldsymbol{E}$ denotes the expected value. We can see that the WD bounds the difference in the expected values. For example, if we compare two posteriors for $\iota$ calculated for two different noise realisations, a WD = 1.0 rad tells us that the difference in the expected value of $\iota$ from the two distributions is less than 1 rad. Beyond optimal transport theory, the WD is a popular metric in machine learning \citep{2017arXiv170107875A}, climate modelling \citep{2022JCli...35.1215P,2023QJRMS.149..843K}, computational biology \citep{GONZALEZDELGADO2023168053} and geophysics \citep{2023GeoRL..5003880M}. A full review of the Wasserstein distance, which may be unfamiliar to some readers, is presented in Appendix \ref{sec:wasserstein} \textcolor{red}{TK: is this too much info on Wasserstein? Maybe just say "its a metric between probability distributions" and move the rest to appendix so as not to interrupt main astrophysics arguments.}

For each parameter in $\boldsymbol{\theta}_{\rm gw}$ we calculate the WD between every pair of posteriors across our $10^3$ noise realisations.

@@

@@

\subsection{Bias by dropping pulsar terms}\label{sec:bias}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/likelihood_iota_new}
	\caption{Variation in the log likelihood returned by the Kalman filter using the Earth terms model (Equation \ref{eq:measuremen_earth}, blue line), the pulsar terms model (Equation \ref{eq:measurement}, orange line), and the Earth terms model for Earth terms only synthetic data (green line), with respect to the inclination $\iota$, over the prior space. All other parameters are fixed at their true injected value from Table \ref{tab:parameters_and_priors}. The coloured dashed vertical lines label the location of the likelihood maxima for their corresponding model. For the Earth-terms model the likelihood maxima does not coincide with the injected value ($\iota = 1.0$). For the pulsar-terms model the location of the maxima and injected value do coincide. 10 noise realisations are plotted but the solutions are completely overlaid.}
	\label{fig:likelihood_surface_iota}
\end{figure}

\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth]{images/large_h_example}
	\caption{} 
	\label{fig:bias_for_large_h}
\end{figure*}

Evident in the preceding discussion is the fact that for some of the $\boldsymbol{\theta}_{\rm gw}$ parameters the inferred value (e.g. the median of the one-dimensional marginalised posterior) is biased away from the true injection value. This effect is most severe for $\iota$ and $h_0$ but also present to a lesser extent in other static parameters such as $\delta,\alpha$ and $\psi$, and additionally to an even smaller extent in $\Phi_0$. In this Section we demonstrate how this bias results from the dropping of the pulsar terms described in Section \ref{sec:parameter_estim}. \newline

for low signal bias is covered by noise

@@

\appendix
\newpage
\section{Kalman recursion equations} \label{sec:kalman}
The linear Kalman filter operate on temporally discrete, noisy measurements $\boldsymbol{Y}_k$ which are related to a set of unobservable discrete system states $\boldsymbol{X}_k$ via a linear transformation:
\begin{equation}
	\boldsymbol{Y}_k = \boldsymbol{H}_k \boldsymbol{X}_k + \boldsymbol{v}_k \label{eq:kalman1}
\end{equation}
where $\boldsymbol{H}_k$ is the measurement matrix or observation model, $\boldsymbol{v}_k$ is a zero-mean Gaussian measurement noise with covariance $\boldsymbol{R}_k$, and the subscript $k$ labels the time-step. The Kalman filter takes the underlying states to evolve according to
\begin{equation}
	\boldsymbol{X}_k = \boldsymbol{F}_k \boldsymbol{X}_{k-1} + \boldsymbol{G}_k \boldsymbol{u}_k + \boldsymbol{w}_k \label{eq:kalman2}
\end{equation}
where $\boldsymbol{F}$ is the system dynamics matrix, $\boldsymbol{G}$ the input or control matrix. and $\boldsymbol{w}_k$ is a zero-mean Gaussian process noise with covariance $\boldsymbol{Q}_k$. \newline

@@

This paper is useful: \citep{KARLSSON2012941} \citep{SEDOGLAVIC2002735}

@@

%\section{References}
%\label{sec:ref_list}
